Synthesizing Symmetric Lenses

### 2018-07-11 papers/SymmetricLenses/main.tex
Comments and polishing

A {\em lens} is a program that can be run both ``front to back'' and ``back
to front,'' allowing data to be transformed in two directions.  Since their
introduction by~\citet{Focal2005-long2}, lenses have been extensively
studied and applied.  Recent work~\cite{optician} has also demonstrated how
techniques from {\em type-directed program synthesis} can be used to
efficiently synthesize a very simple class of lenses---so-called {\em
  bijective lenses} over string data---given a pair of types (regular
expressions) and a small number of examples.

@@

We describe an implementation of simple symmetric lenses and our synthesis
procedure as extensions to the Boomerang language~\cite{Boomerang}. We
evaluate
its performance on 48 benchmark examples drawn from FlashFill, Augeus,
and an online collection of electronic file formats. We find this implementation is
able to synthesize all the lenses of the benchmark suite in under 30 seconds.
\end{abstract}

@@

\noindent
In a big-data world, similar data is often stored in a variety of
formats. For instance, electronic calendars come in iCalendar,
vCalendar and h-event formats~\cite{calendar-formats}; US taxation data can be
transmitted via Tax XML (used by the US government for e-filing) or TXF (used by
TurboTax)~\cite{tax-formats}; and sewing machines can read embroidery files
transmitted as PES formats, DST formats or several
others~\cite{embroidery-formats}.

One way to interconvert between these varied data formats is to use a
\emph{lens}~\cite{Focal2005}---a bi-directional
program that can transform data represented in a format $S$ to a format $T$,
and vice versa.  There are many different kinds of lenses, but well-defined lens
languages typically provide guarantees that help ensure data is
not lost or corrupted as it is transformed back and forth between different representations.

While lenses facilitate writing principled data
transformations, \textit{synthesizing} lenses 
can make the task even easier.  Lens languages are a ripe target for
synthesis because lens languages, like many domain-specific languages,
reduce the space of possible programs, making search-based synthesis easier. 
%This feature can be a negative, if the desired transformation
%falls outside of the scope provided by the language, but can also be positive, as complex
%properties (like invertability) can be enforced automatically.
%Another advantage of reducing the
%set of programs is that doing so can facilitate program synthesis, which is essentially
%a search through the space of possible computations that can be expressed as a lens.
Recent research~\cite{optician, maina+:quotient-synthesis} has exploited this
observation by demonstrating that it is possible to synthesize relatively large,
complex lenses from specifications. In particular, given types that describe
source and target formats and optional examples of corresponding data, the
bijective synthesis algorithm presented in these 
papers will find bijective lenses to convert between data stored
% ~\cite{optician}
in different formats,
such as linux configuration files from different systems that represent the same data in different ways.  
Often, these algorithms can find an appropriate lens
in less than a second without requiring any examples.
%, with just types describing the source and target formats.
% \dpw{I would cite the ICFP paper (cite string:
%   maina+:quotient-synthesis) as to appear but I'm not sure how to do that
%   without possibly violating the spirit of double-blind reviewing. ie, if it
%   hasn't appeared yet, how could we know the content if we aren't the authors?
%   Mentioning it in the 3rd person does not seem to suffice ...} \ksf{The ICFP
%   2018 website lists the paper; I think it is okay to cite it. The reviewer can
%   assume we contacted the authors for a copy. We could ask Stephanie for a
%   ruling if we want to be sure. One of the principles of double blind as adopted
%   by SIGPLAN is not to reduce the quality of the paper; omitting this citation
%   would have that affect.}\bcp{+1}\saz{Stephanie says we should cite it in the
%   third person, even if that's a bit awkward, but otherwise citing it is OK.}
One reason that bijective lens synthesis is so effective, even relative to successful
synthesis projects in other domains, is that there are not very
many bijections between related data formats (unless the
same kind of data appears in the two formats many times). 
Hence, if the synthesis algorithm can
find any bijection at all, then it is very likely to find the one the
user wants.

@@

With simple symmetric lenses and our SRE-based likeliness measure in
hand, we develop a new algorithm for synthesizing likely lenses.
At its core, the algorithm performs a type-directed search between the 
descriptions of the source and target formats, measuring success using
the likeliness measure.
However, regular expressions enjoy a rich algebraic structure, meaning
there are infinitely many regular expressions equivalent to the
original pair, and the lens returned by a type-directed search depends
upon the chosen representation. 
To tame this complexity, we divide the synthesis algorithm into two communicating
search procedures (Figure~\ref{fig:high-level-algorithm}), following
earlier work~\cite{maina+:quotient-synthesis}.
The first algorithm, \Expand, uses rewriting rules to iteratively
propose new pairs of stochastic regular expressions equivalent to the
original pair.
The second algorithm, \GreedySynth, takes a pair of such SREs as input
and uses a greedy, type-directed algoithm to find a simple symmetric
lens between them, returning the lens and its likelihood score
to \Expand.  The algorithm terminates when the likilihood score of the
best lens found so far compared with 
the complexity of newly generated SREs 
suggests that further searching is unlikely to be fruitful.

@@

In summary, our main contributions are:
\begin{itemize}
\item We define simple symmetric lenses, which have
  the useful property that they do not maintain any hidden state
  (\S\ref{sec:overview}).  
  We give a language for defining such lenses and describe how to
  calculate their likelihoods (\S\ref{sec:ssl}). 
\item  We show that simple symmetric lenses
  fall between asymmetric lenses~\cite{Focal2005-long2} and full
  symmetric lenses~\cite{symmetric-lenses} in expressiveness
  (\S\ref{sec:related}).  We further show that the class of simple
  lenses can be characterized semantically as the subset of full
  symmetric lenses for which a simple ``forgetfulness'' property holds
  (\S\ref{sec:relationship}).
\item We show how to synthesize simple symmetric lenses (including all
  asymmetric lenses as a special case), using a novel application of stochastic
  regular expressions to guide the search process (\S\ref{sec:synthesis}). 
  To support that effort, we show that the star semi-ring equivalences on regular
  expressions remain valid when applied to stochastic regular
  expressions, preserving the probability distributions as well as the languages
  (\S\ref{sec:sre}).
\item We extend Boomerang with simple symmetric lenses,
  showing that such lenses can integrate smoothly with 
  other lens extensions such as quotient lenses and matching
  lenses.  We evaluate our
  implementation on a set of 48 lens synthesis benchmarks drawn from
  data cleaning, view maintenance, 
  and file synchronization tasks.  We demonstrate that we can
  synthesize simple symmetric lenses for all of the benchmarks in
  under 30 seconds (\S\ref{sec:evaluation}). 
\end{itemize}
We finish the paper with a discussion of related work (\S\ref{sec:related}) and
our concluding thoughts (\S\ref{sec:conc}).

@@

\subsection{Synthesizing Symmetric Lenses}

@@

Now that we have a way to calculate likelihood, we need a way to search for
likely lenses. Previous work on synthesis~\cite{augustsson-2004,osera+:pldi15,feser-pldi-2015,frankle+:popl16} has shown that we can use
types to dramatically restrict the search space and improve the effectiveness of
synthesis. In our setting, types are pairs of stochastic regular expressions,
which come with a rich algebraic structure that makes each regular expression
semantically equal to infinitely many others. To tame that search space and
leverage the benefits of type-directed synthesis, we split our algorithm into
two communicating synthesizers following earlier
work~\cite{maina+:quotient-synthesis}. The first synthesizer, \Expand, navigates
the space of semantically equivalent regular expressions by applying rewrite
rules that preserve both the semantics and the probability distributions of
those stochastic regular expressions. Our algorithm ranks pairs of stochastic
regular expressions by the number of rewrite rule applications required to
obtain the pair from the stochastic regular expressions given as input. It
passes the pairs off to the second synthesizer, \GreedySynth, in rank order with
the smallest first.

@@

\section{Stochastic Regular Expressions}
\label{sec:sre}
To characterize the likely lenses, we must compute the expected number of bits
needed to recover a string in one data source from a synchronized string in the
other data source. To do this, we must first develop a probabilistic language
model for our language. We do this with \emph{stochastic regular expressions},
regular expressions annotated with probability information. With this, we can
jointly express a language and a probability distribution over that language. In
the following grammar, $\String$ ranges over strings, and $\Probability$ ranges
over real numbers between 0 and 1, exclusive.

$$\Regex{},\RegexAlt{} \GEq{}  \String  \; | \; \emptyset \; | \; \PRegexStar{\Regex}{p} \; | \; \RegexConcat{\Regex_1}{\Regex_2} \; | \; \PRegexOr{\Regex_1}{\Regex_2}{p}$$

@@

\begin{figure}
  \begin{tabular}{@{}r@{\hspace{1em}}c@{\hspace{1em}}l@{}r@{}}
    \PRegexOr{\Regex}{\emptyset}{1} & $\SSREquiv$ & \Regex{} & \OrIdentityRule{} \\
    $\RegexConcat{\Regex}{\emptyset}$ & $\SSREquiv$ & $\emptyset$ & \EmptyProjectionRightRule{} \\
    $\RegexConcat{\emptyset}{\Regex}$ & $\SSREquiv$ & $\emptyset$ & \EmptyProjectionLeftRule{} \\
    \RegexConcat{(\RegexConcat{\Regex{}}{\Regex'})}{\Regex''} & $\SSREquiv$ & \RegexConcat{\Regex{}}{(\RegexConcat{\Regex'}{\Regex''})} & \ConcatAssocRule{}  \\
    \PRegexOr{(\PRegexOr{\Regex}{\Regex'}{\Probability_1})}{\Regex''}{\Probability_2} & $\SSREquiv$ & \PRegexOr{\Regex}{(\PRegexOr{\Regex'}{\Regex''}{\frac{(1-\Probability_1)\Probability_2}{1 - \Probability_1\Probability_2}})}{\Probability_1\Probability_2} & \OrAssociativityRule{}  \\
    \PRegexOr{\Regex{}}{\RegexAlt{}}{\Probability} & $\SSREquiv$ & \PRegexOr{\RegexAlt{}}{\Regex{}}{1-\Probability} & \OrCommutativityRule{}\\
    \RegexConcat{\Regex{}}{(\PRegexOr{\Regex{}'}{\Regex{}''}{\Probability})} & $\SSREquiv$ & \PRegexOr{(\RegexConcat{\Regex{}}{\Regex{}'})}{(\RegexConcat{\Regex{}}{\Regex{}''})}{\Probability} & \DistributivityLeftRule{} \\
    \RegexConcat{(\PRegexOr{\Regex{}'}{\Regex{}''}{\Probability})}{\Regex{}} & $\SSREquiv$ & \PRegexOr{(\RegexConcat{\Regex{}'}{\Regex{}})}{(\RegexConcat{\Regex{}''}{\Regex{}})}{\Probability} & \DistributivityRightRule{} \\
    \RegexConcat{\EmptyString{}}{\Regex{}} & $\SSREquiv$ & \Regex{} & \ConcatIdentityLeftRule{} \\
    \RegexConcat{\Regex{}}{\EmptyString{}} & $\SSREquiv$ & \Regex{} & \ConcatIdentityRightRule{} \\
    \PRegexStar{\Regex{}}{\Probability} & $\SSREquiv$ & \PRegexOr{\EmptyString{}}{(\RegexConcat{\Regex{}}{\PRegexStar{\Regex{}}{\Probability}})}{1-\Probability} & \UnrollstarLeftRule{} \\
    \PRegexStar{\Regex{}}{\Probability} & $\SSREquiv$ & \PRegexOr{\EmptyString{}}{(\RegexConcat{\PRegexStar{{\Regex{}}}{\Probability}}{\Regex{}})}{1-\Probability} & \UnrollstarRightRule{} 
  \end{tabular}
  \caption{Stochastic Regular Expression Star-Semiring Equivalence}
  \label{fig:ssr-sre}
\end{figure}

@@

% begin synthesis
\section{Synthesis}
\label{sec:synthesis}
Algorithm~\ref{alg:synth-sym-lens} presents our synthesis algorithm at a high
level of abstraction. The input regular expressions are first converted into
stochastic regular expressions with \ToStochastic.  
%Once that conversion has
%taken place, the $\RXSearch$ algorithm executes. This algorithm itself is
%defined via five functions: (1) $\PCF{\PQ.Create}$, (2) $\PCF{\PQ.Pop}$, (3)
%$\PCF{Expand}$, (4) $\PCF{\PQ.Push}$, and (5) $\PCF{\Continue}$. First,
This pair of SREs is used to initialize a priority queue ($pq$).
Next, $\PCF{SynthSymLens}$
enters a loop that searches for likely lenses. The loop terminates when the
algorithm believes it is unlikely to find a better lens than the best one it has
found so far (a termination condition defined by $\PCF{\Continue}$). Within each
iteration of the loop, it:
\begin{itemize}
\item pops the next class ($S$, $T$) of lenses to
search off of the priority queue ($\PCF{\PQ.Pop}$),
\item executes $\GreedySynth$ to find a best lens in that class if
one exists ($\LC$),
using the examples $exs$ to filter out potential lenses that do not satisfy
the specification,
\item replaces $best$ with $\LC$, if $\LC$ has lower cost according
to our information theoretic metric, and
\item adds the SREs derived from rewriting $S$ and $T$ ($\Expand(S,T)$)
  to the priority queue.  The priority of a SRE pair is the
  number of rewrites needed to derive the pair from the originals.
\end{itemize}
When the loop terminates, the search returns the globally best lens found
($best$).  Each subroutine of this algorithm will be explained in
further depth in the following subsections.

@@

The algorithm for converting a stochastic regular expressions into its
corresponding stochastic DNF regular expressions, written
$\ToDNFRegexOf{\Regex}$, is defined below.
\[
  \begin{array}{rcl}
    \ToDNFRegexOf{\String} & = & \DNFOf{(\SequenceOf{\String},1)}\\
    \ToDNFRegexOf{\emptyset} & = & \DNFOf{}\\
    \ToDNFRegexOf{(\PRegexStar{\Regex}{\Probability})} & = & \AtomToDNFOf{\PRegexStar{(\ToDNFRegexOf{\Regex})}{\Probability}}\\
    \ToDNFRegexOf{(\RegexConcat{\Regex_1}{\Regex_2})} & = & \ToDNFRegexOf{\Regex_1} \ConcatDNF \ToDNFRegexOf{\Regex_2}\\
    \ToDNFRegexOf{(\PRegexOr{\Regex_1}{\Regex_2}{\Probability})} & = & \ToDNFRegexOf{\Regex_1} \OrDNF_{\Probability} \ToDNFRegexOf{\Regex_2}\\
  \end{array}
\]
This conversion relies on operators defined in
Figure~\ref{fig:dnf-regex-functions}. We have proved that this conversion
respects languages and probability distributions.

\begin{theorem}
  $\ProbabilityOf{\Regex}{\String} = \ProbabilityOf{\ToDNFRegex
    \Regex}{\String}$ and $\LanguageOf{\Regex} =
  \LanguageOf{\ToDNFRegexOf{\Regex}}$
\end{theorem}

@@

\subsection{The Effects of Our Heuristics}
\label{subsec:effect-of-heuristics}
We evaluate the usefulness of (1) our information theoretic metric, and (2) our
termination heuristic.  To this end, we run our program in several different modes:\\
\begin{tabulary}{\linewidth}{rL}
  \AnyOpt{}: & We synthesize lenses without use of the information
               theoretic preference metric.  Once \GreedySynth finds a lens, the
  synthesis procedure returns that lens.\\
  \FLOpt{}: &  The first time \GreedySynth finds a lens, the synthesis procedure
              returns that lens. However, \GreedySynth uses the information
              theoretic preference metric to choose a lens.\\
  \CCOpt{}: & We replace our information theoretic cost metric with one where
              the cost of the lens is the number of disonnects plus the number
              of merges.\\
  \ZTOpt{}: & \ZTOpt{} runs the algorithm with the termination parameter
              universally set to 0 (default). \\
  \TFTOpt{}: & \TFTOpt{} runs the algorithm with the
               termination parameter universally set to 25. \\
  \NTFTOpt{}: & \NTFTOpt{} runs the algorithm with the
                termination parameter universally set to -25. \\
\end{tabulary}

@@

We experimented with the \CCOpt{} mode to determine whether the complexity of the information theoretic measure was really needed. Related work on string transformations has often used simpler measures such as ``avoid constants'' which align with, but are simpler than our measures. The \CCOpt{} mode is an example of such a simple measure---it operates by counting disconnects, which put a complete stop to information transfer, like constants, and merges, which eliminate the information in a union. Figure~\ref{fig:metric} summarizes the result of this experiment. We find that
our metrics are very important for finding the correct lens. Our
information-theoretic metric is critical for finding the correct lens. Indeed,
only 10 of our benchmarks succeeded when running in \CCOpt{} mode.

@@

\begin{theorem}
  Let $\Lens$ be an asymmetric lens. $\Lens$ is also a simple symmetric lens,
  where
  \begin{center}
    \begin{tabular}{rcl}
      $\Lens.\CreateROf{x}$ & $=$ & $\Lens.get \App x$\\
      $\Lens.\CreateLOf{y}$ & $=$ & $\Lens.create \App y$\\
      $\Lens.\PutROf{x}{y}$ & $=$ & $\Lens.get \App x$\\
      $\Lens.\PutLOf{y}{x}$ & $=$ & $\Lens.put \App y \App x$
    \end{tabular}
  \end{center}
\end{theorem}

@@

All pragmatic synthesis algorithms use heuristics of one kind or another.  One of the
goals of the current paper is to try to ground those heuristics in a broader theory,
information theory,
with the hope that this theory may inform future design decisions and help us understand
heuristics crafted in other tools and possibly in other domains.  For instance, we
speculate that some of the heuristics used in
FlashFill (to take one well-documented example) may be connected to some of the principles
laid out here.  For instance, Flash Fill prioritizes the substring constructor over the
constant string constructor when ranking possible programs. Such a choice is
consistent with our information-theoretic viewpoint as the constant function
throws away a great deal of information about the source string being
transformed. Likewise, Flash Fill prefers ``wider'' character classes over
``narrower'' ones. Again such a choice is implied by information theory -- the
wider class preserves more information during translation.  More broadly, we
hope our information theoretic analysis provides a principle that can be
used to understand the heuristic choices made in related work.

>>>>

{\em Lenses} are programs that can be run both ``front to back'' and ``back
to front,'' allowing data to be transformed in two directions.  Since their
introduction by~\citet{Focal2005-long2}, lenses have been extensively
studied and applied.  Recent work~\cite{optician} has also demonstrated how
techniques from {\em type-directed program synthesis} can be used to
efficiently synthesize a very simple class of lenses---so-called {\em
  bijective lenses} over string data---given a pair of types (regular
expressions) and a small number of examples.

@@

We describe an implementation of simple symmetric lenses and our synthesis
procedure as extensions to the Boomerang language~\cite{Boomerang}. We
evaluate
its performance on 48 benchmark examples drawn from FlashFill~\bcp{cite}, Augeus~\bcp{cite},
and an online collection of electronic file formats. Our
implementation can synthesize all these lenses in under
30 seconds. 
\end{abstract}

@@

\noindent
In the big-data world, similar information is often stored or transmitted in
different formats. For instance, electronic calendars come in iCalendar,
vCalendar, and h-event formats~\cite{calendar-formats}; US taxation data can
be transmitted via Tax XML (used by the US government for e-filing) or TXF
(used by TurboTax)~\cite{tax-formats}; and sewing machines can use
embroidery files in formats with names like PES, DST, and several
others~\cite{embroidery-formats}.  \bcp{It would be nice to be able to give
  a quick example of a truly symmetric case---do any of these examples work
  for that?}

One convenient way to interconvert between these varied data formats is to
use a \emph{lens}~\cite{Focal2005}---a bidirectional program that can
transform data represented in a format $S$ to a format $T$, and vice versa.
offering ``round tripping'' guarantees that help ensure data is not lost or
corrupted as it is transformed back and forth between different
representations.

While domain-specific languages for writing lenses can facilitate building
principled data transformations, \textit{synthesizing} lenses 
can make the task even easier.  Indeed, lens languages offer excellent
targets for synthesis, since their specialized, sub-Turing-complete syntax
and expressive type systems drastically
reduce the space of possible programs, making search-based synthesis easier. 
%
Recent work~\cite{optician} has exploited this
observation, demonstrating that it is feasible to synthesize quite complex
examples, albeit only for a simple form of lenses---so-called bijective
ones. Specifically, given a pair of types describing the intended
source and target formats plus a small number of examples of corresponding
data, the bijective synthesis algorithm developed in these two
papers will find a bijective lens that converts between data stored
in different formats,
such as linux configuration files from different systems that represent the same data in different ways.  
Often, these algorithms can find an appropriate lens
in less than a second without requiring any examples.
%, with just types describing the source and target formats.
% \dpw{I would cite the ICFP paper (cite string:
%   maina+:quotient-synthesis) as to appear but I'm not sure how to do that
%   without possibly violating the spirit of double-blind reviewing. ie, if it
%   hasn't appeared yet, how could we know the content if we aren't the authors?
%   Mentioning it in the 3rd person does not seem to suffice ...} \ksf{The ICFP
%   2018 website lists the paper; I think it is okay to cite it. The reviewer can
%   assume we contacted the authors for a copy. We could ask Stephanie for a
%   ruling if we want to be sure. One of the principles of double blind as adopted
%   by SIGPLAN is not to reduce the quality of the paper; omitting this citation
%   would have that affect.}\bcp{+1}\saz{Stephanie says we should cite it in the
%   third person, even if that's a bit awkward, but otherwise citing it is OK.}
One reason that bijective lens synthesis is so effective, even relative to successful
synthesis projects in other domains, is that there are not very
many bijections between related data formats (unless the
same kind of data appears in the two formats many times). 
Hence, if the synthesis algorithm can
find any bijection at all, then it is very likely to find the one the
user wants.

@@

With simple symmetric lenses and our SRE-based likeliness measure in
hand, we develop a new algorithm for synthesizing likely lenses.
At its core, the algorithm performs a type-directed search between the 
descriptions of the source and target formats, measuring success using
the likeliness measure.
However, regular expressions enjoy a rich algebraic structure, meaning
there are infinitely many regular expressions equivalent to the
original pair, and the lens returned by a type-directed search depends
upon the chosen representation. 
To tame this complexity, we divide the synthesis algorithm into two communicating
search procedures (Figure~\ref{fig:high-level-algorithm}), following
earlier work~\cite{maina+:quotient-synthesis}\bcp{Is this the right citation?}.
The first algorithm, \Expand, uses rewriting rules to iteratively
propose new pairs of stochastic regular expressions equivalent to the
original pair.
The second algorithm, \GreedySynth, takes a pair of such SREs as input
and uses a greedy, type-directed algoithm to find a simple symmetric
lens between them, returning the lens and its likelihood score
to \Expand.  The algorithm terminates when the likilihood score of the
best lens found so far compared with 
the complexity of newly generated SREs 
suggests that further searching is unlikely to be fruitful.

@@

In summary, our contributions are:\bcp{The section references do not proceed
  in order, so this doesn't work very well as an outline of the paper.
  Should we try to reorder the contributions?  Write a separate outline?
  Live with it as-is?}
\begin{itemize}
\item We define {\em simple symmetric lenses}, with the
  the useful property that they do not maintain any hidden state
  (\S\ref{sec:overview}).  
  We present a collection of combinators for defining these lenses and
  describe how to 
  calculate their likelihoods (\S\ref{sec:ssl}). 
\item  We show that simple symmetric lenses
  fall between asymmetric lenses~\cite{Focal2005-long2} and full
  symmetric lenses~\cite{symmetric-lenses} in expressiveness
  (\S\ref{sec:related}).  We further show that the class of simple
  lenses can be characterized semantically as the subset of full
  symmetric lenses for which a simple ``forgetfulness'' property holds.
\item We show how to synthesize simple symmetric lenses (including all
  asymmetric lenses as a special case), using a novel application of stochastic
  regular expressions to guide the search process (\S\ref{sec:synthesis}). 
  To support that effort, we show that the star semi-ring equivalences on
  regular 
  expressions remain valid when applied to stochastic regular
  expressions, preserving the probability distributions as well as the languages
  (\S\ref{sec:sre}).
\item We extend the Boomerang implementation with simple symmetric lenses,
  demonstrating that they integrate smoothly with 
  other lens extensions such as quotient lenses~\bcp{citation} and matching
  lenses~\bcp{citation}.  We evaluate our
  implementation on a set of 48 lens synthesis benchmarks drawn from
  data cleaning, view maintenance, 
  and file synchronization tasks.  We demonstrate that we can
  synthesize simple symmetric lenses for all of the benchmarks in
  under 30 seconds (\S\ref{sec:evaluation}). 
\end{itemize}
We close with a discussion of related work (\S\ref{sec:related}) and
concluding thoughts (\S\ref{sec:conc}).

@@

\subsection{Synthesizing Simple Symmetric Lenses}

@@

Now that we have a way to calculate likelihood, we need a way to search for
likely lenses. Previous work on synthesis~\cite{augustsson-2004,osera+:pldi15,feser-pldi-2015,frankle+:popl16} has shown that we can use
types to dramatically restrict the search space and improve the effectiveness of
synthesis. In our setting, types are pairs of stochastic regular expressions,
which come with a rich algebraic structure that makes each regular expression
semantically equal to infinitely many others. To tame that search space and
leverage the benefits of type-directed synthesis, we split our algorithm into
two communicating synthesizers following earlier
work~\cite{maina+:quotient-synthesis}\bcp{Some of this text seems to be
  identical to text in the intro.  That will annoy careful readers.}. The
first synthesizer, \Expand, navigates 
the space of semantically equivalent regular expressions by applying rewrite
rules that preserve both the semantics and the probability distributions of
those stochastic regular expressions. Our algorithm ranks pairs of stochastic
regular expressions by the number of rewrite rule applications required to
obtain the pair from the stochastic regular expressions given as input. It
passes the pairs off to the second synthesizer, \GreedySynth, in rank order with
the smallest first.

@@

\section{Stochastic Regular Expressions}
\label{sec:sre}
To characterize the likely lenses, we must compute the expected number of bits
needed to recover a string in one data source from a synchronized string in the
other data source. To do this, we must first develop a probabilistic language
model for our language. We do this with \emph{stochastic regular expressions},
regular expressions annotated with probability information. With this, we can
jointly express a language and a probability distribution over that language. In
the following grammar, $\String$ ranges over strings, and $\Probability$ ranges
over real numbers between 0 and 1, exclusive.
$$\Regex{},\RegexAlt{} \GEq{}  \String  \; | \; \emptyset \; | \; \PRegexStar{\Regex}{p} \; | \; \RegexConcat{\Regex_1}{\Regex_2} \; | \; \PRegexOr{\Regex_1}{\Regex_2}{p}$$

@@

\begin{figure}
  \begin{tabular}{@{}r@{\hspace{1em}}c@{\hspace{1em}}l@{}r@{}}
    \PRegexOr{\Regex}{\emptyset}{1} & $\SSREquiv$ & \Regex{} & \OrIdentityRule{} \\
    $\RegexConcat{\Regex}{\emptyset}$ & $\SSREquiv$ & $\emptyset$ & \EmptyProjectionRightRule{} \\
    $\RegexConcat{\emptyset}{\Regex}$ & $\SSREquiv$ & $\emptyset$ & \EmptyProjectionLeftRule{} \\
    \RegexConcat{(\RegexConcat{\Regex{}}{\Regex'})}{\Regex''} & $\SSREquiv$ & \RegexConcat{\Regex{}}{(\RegexConcat{\Regex'}{\Regex''})} & \ConcatAssocRule{}  \\
    \PRegexOr{(\PRegexOr{\Regex}{\Regex'}{\Probability_1})}{\Regex''}{\Probability_2} & $\SSREquiv$ & \PRegexOr{\Regex}{(\PRegexOr{\Regex'}{\Regex''}{\frac{(1-\Probability_1)\Probability_2}{1 - \Probability_1\Probability_2}})}{\Probability_1\Probability_2} & \OrAssociativityRule{}  \\
    \PRegexOr{\Regex{}}{\RegexAlt{}}{\Probability} & $\SSREquiv$ & \PRegexOr{\RegexAlt{}}{\Regex{}}{1-\Probability} & \OrCommutativityRule{}\\
    \RegexConcat{\Regex{}}{(\PRegexOr{\Regex{}'}{\Regex{}''}{\Probability})} & $\SSREquiv$ & \PRegexOr{(\RegexConcat{\Regex{}}{\Regex{}'})}{(\RegexConcat{\Regex{}}{\Regex{}''})}{\Probability} & \DistributivityLeftRule{} \\
    \RegexConcat{(\PRegexOr{\Regex{}'}{\Regex{}''}{\Probability})}{\Regex{}} & $\SSREquiv$ & \PRegexOr{(\RegexConcat{\Regex{}'}{\Regex{}})}{(\RegexConcat{\Regex{}''}{\Regex{}})}{\Probability} & \DistributivityRightRule{} \\
    \RegexConcat{\EmptyString{}}{\Regex{}} & $\SSREquiv$ & \Regex{} & \ConcatIdentityLeftRule{} \\
    \RegexConcat{\Regex{}}{\EmptyString{}} & $\SSREquiv$ & \Regex{} & \ConcatIdentityRightRule{} \\
    \PRegexStar{\Regex{}}{\Probability} & $\SSREquiv$ & \PRegexOr{\EmptyString{}}{(\RegexConcat{\Regex{}}{\PRegexStar{\Regex{}}{\Probability}})}{1-\Probability} & \UnrollstarLeftRule{} \\
    \PRegexStar{\Regex{}}{\Probability} & $\SSREquiv$ & \PRegexOr{\EmptyString{}}{(\RegexConcat{\PRegexStar{{\Regex{}}}{\Probability}}{\Regex{}})}{1-\Probability} & \UnrollstarRightRule{} 
  \end{tabular}
  \caption{Stochastic Regular Expression Star-Semiring Equivalence\bcp{Can
      we make the skips between the separate sections of the definition a
      bit larger, and increase the margin on the left?  It looks (even)
      denser than it is, at the moment.}}
  \label{fig:ssr-sre}
\end{figure}

@@

% begin synthesis
\section{Synthesis}
\label{sec:synthesis}
Algorithm~\ref{alg:synth-sym-lens} presents our synthesis algorithm at a high
level of abstraction. The input regular expressions are first converted into
stochastic regular expressions with \ToStochastic.  
%Once that conversion has
%taken place, the $\RXSearch$ algorithm executes. This algorithm itself is
%defined via five functions: (1) $\PCF{\PQ.Create}$, (2) $\PCF{\PQ.Pop}$, (3)
%$\PCF{Expand}$, (4) $\PCF{\PQ.Push}$, and (5) $\PCF{\Continue}$. First,
This pair of SREs is used to initialize a priority queue ($pq$).
Next, $\PCF{SynthSymLens}$
enters a loop that searches for likely lenses. The loop terminates when the
algorithm believes it is unlikely to find a better lens than the best one it has
found so far (a termination condition defined by $\PCF{\Continue}$). Within each
iteration of the loop, it:
\begin{itemize}
\item pops the next class ($S$, $T$) of lenses to
search off of the priority queue ($\PCF{\PQ.Pop}$),
\item executes $\GreedySynth$ to find a best lens in that class if
one exists ($\LC$),
using the examples $exs$ to filter out potential lenses that do not satisfy
the specification,
\item replaces $best$ with $\LC$, if $\LC$ has lower cost according
to our information-theoretic metric, and
\item adds the SREs derived from rewriting $S$ and $T$ ($\Expand(S,T)$)
  to the priority queue.  The priority of a SRE pair is the
  number of rewrites needed to derive the pair from the originals.
\end{itemize}
When the loop terminates, the search returns the globally best lens found
($best$).  Each subroutine of this algorithm will be explained in
further depth in the following subsections.

@@

The algorithm for converting a stochastic regular expressions into its
corresponding stochastic DNF regular expressions, written
$\ToDNFRegexOf{\Regex}$, is defined below.\bcp{Could put this in two columns
to save space.}
\[
  \begin{array}{rcl}
    \ToDNFRegexOf{\String} & = & \DNFOf{(\SequenceOf{\String},1)}\\
    \ToDNFRegexOf{\emptyset} & = & \DNFOf{}\\
    \ToDNFRegexOf{(\PRegexStar{\Regex}{\Probability})} & = & \AtomToDNFOf{\PRegexStar{(\ToDNFRegexOf{\Regex})}{\Probability}}\\
    \ToDNFRegexOf{(\RegexConcat{\Regex_1}{\Regex_2})} & = & \ToDNFRegexOf{\Regex_1} \ConcatDNF \ToDNFRegexOf{\Regex_2}\\
    \ToDNFRegexOf{(\PRegexOr{\Regex_1}{\Regex_2}{\Probability})} & = & \ToDNFRegexOf{\Regex_1} \OrDNF_{\Probability} \ToDNFRegexOf{\Regex_2}\\
  \end{array}
\]
This conversion relies on operators defined in
Figure~\ref{fig:dnf-regex-functions}. We have proved that this conversion
respects languages and probability distributions.

\begin{theorem}
  $\ProbabilityOf{\Regex}{\String} = \ProbabilityOf{\ToDNFRegex
    \Regex}{\String}$ and $\LanguageOf{\Regex} =
  \LanguageOf{\ToDNFRegexOf{\Regex}}$.
\end{theorem}

@@

\subsection{The Effects of Our Heuristics}
\label{subsec:effect-of-heuristics}
We evaluate the usefulness of (1) our information-theoretic metric, and (2) our
termination heuristic.  To this end, we run our program in several different modes:\\
\begin{tabulary}{\linewidth}{rL}
  \AnyOpt{}: & We synthesize lenses without use of the information-theoretic
  preference metric.  Once \GreedySynth finds a lens, the 
  synthesis procedure returns that lens.\\
  \FLOpt{}: &  The first time \GreedySynth finds a lens, the synthesis procedure
              returns that lens. However, \GreedySynth uses the
              information-theoretic preference metric to choose a lens.\\ 
  \CCOpt{}: & We replace our information-theoretic cost metric with one where
              the cost of the lens is the number of disonnects plus the number
              of merges.\\
  \ZTOpt{}: & \ZTOpt{} runs the algorithm with the termination parameter
              universally set to 0 (default). \\
  \TFTOpt{}: & \TFTOpt{} runs the algorithm with the
               termination parameter universally set to 25. \\
  \NTFTOpt{}: & \NTFTOpt{} runs the algorithm with the
                termination parameter universally set to -25. \\
\end{tabulary}

@@

We experimented with the \CCOpt{} mode to determine whether the complexity of the information-theoretic measure was really needed. Related work on string transformations has often used simpler measures such as ``avoid constants'' which align with, but are simpler than our measures. The \CCOpt{} mode is an example of such a simple measure---it operates by counting disconnects, which put a complete stop to information transfer, like constants, and merges, which eliminate the information in a union. Figure~\ref{fig:metric} summarizes the result of this experiment. We find that
our metrics are very important for finding the correct lens. Our
information-theoretic metric is critical for finding the correct lens. Indeed,
only 10 of our benchmarks succeeded when running in \CCOpt{} mode.

@@

\begin{theorem}
  Let $\Lens$ be an asymmetric lens. $\Lens$ is also a simple symmetric lens,
  where:\bcp{This could be set in two columns.}
  \begin{center}
    \begin{tabular}{rcl}
      $\Lens.\CreateROf{x}$ & $=$ & $\Lens.get \App x$\\
      $\Lens.\CreateLOf{y}$ & $=$ & $\Lens.create \App y$\\
      $\Lens.\PutROf{x}{y}$ & $=$ & $\Lens.get \App x$\\
      $\Lens.\PutLOf{y}{x}$ & $=$ & $\Lens.put \App y \App x$
    \end{tabular}
  \end{center}
\end{theorem}

@@

All pragmatic synthesis algorithms use heuristics of one kind or another.  One of the
goals of the current paper is to try to ground those heuristics in a broader theory,
information theory,
with the hope that this theory may inform future design decisions and help us understand
heuristics crafted in other tools and possibly in other domains.  For instance, we
speculate that some of the heuristics used in
FlashFill (to take one well-documented example) may be connected to some of the principles
laid out here.  For instance, Flash Fill prioritizes the substring constructor over the
constant string constructor when ranking possible programs. Such a choice is
consistent with our information-theoretic viewpoint as the constant function
throws away a great deal of information about the source string being
transformed. Likewise, Flash Fill prefers ``wider'' character classes over
``narrower'' ones. Again such a choice is implied by information theory -- the
wider class preserves more information during translation.  More broadly, we
hope our information-theoretic analysis provides a basis for
understanding the heuristic choices made in related work.
